### 2019-08-22 submission/manuscript.tex
Making changes in response to Pat's edits

Proton pump inhibitor (PPI) use has been associated with microbiota
alterations and susceptibility to \emph{Clostridium difficile}
infections (CDIs) in humans. We assessed how PPI treatment alters the
fecal microbiota and whether treatment promotes CDIs in a mouse model.
Mice receiving a PPI treatment were gavaged with 40 mg/kg of omeprazole
during a 7-day pretreatment phase, the day of \emph{C. difficile}
challenge, and the following 9 days. We found that mice treated with
omeprazole were not colonized by \emph{C. difficile}. When omeprazole
treatment was combined with a single clindamycin treatment, one cage of
mice remained resistant to \emph{C. difficile} colonization, while the
other cage was colonized. Treating mice with only clindamycin followed
by challenge resulted in \emph{C. difficile} colonization. 16S rRNA gene
sequencing analysis revealed that omeprazole had minimal impact on the
structure of the murine microbiota throughout the 16 days of omeprazole
exposure. These results suggest omeprazole treatment alone is not
sufficient to disrupt microbiota resistance to \emph{C. difficle}
infection in mice that are normally resistant in the absence of
antibiotic treatment.

@@

Antibiotics are the primary risk factor for \emph{Clostridium difficile}
infections (CDIs), but other factors may also increase a person's risk.
In epidemiological studies, PPI use has been associated with CDI
incidence and recurrence. Proton pump inhibitors (PPIs) have also been
associated with alterations in the human intestinal microbiota in
observational and interventional studies. We evaluated the effects of
the PPI omeprazole on the structure of the murine intestinal microbiota
and its ability to disrupt colonization resistance to \emph{C.
difficile}. We found omeprazole treatment had minimal impact on the
murine fecal microbiota and did not promote \emph{C. difficile}
colonization. Further studies are needed to determine whether other
factors such as the composition of the starting bacterial community,
comorbidities, and use of additional medications contribute to the
association between PPIs and CDIs seen in humans or whether aspects of
murine physiology may limit its utility to test these types of
hypotheses.

@@

Antibiotics have a large impact on the intestinal microbiome and are a
primary risk factor for developing \emph{Clostridium difficile}
infections (CDIs) (1). It is less clear whether other human medications
that impact the microbiota also influence \emph{C. difficile}
colonization resistance. Multiple epidemiological studies have suggested
an association between proton pump inhibitor (PPI) use and incidence or
recurrence of CDIs (2--5). There have also been a number of large cohort
studies and interventional clinical trials that demonstrated specific
alterations in the intestinal microbiome were associated with PPI use
(4, 6). PPI-associated microbiota changes have been attributed to the
ability of PPIs to increase stomach acid pH which may promote the
survival of oral and pathogenic bacteria (4, 6). In human fecal samples,
PPI use results in increases in Enterococcaceae, Lactobacillaceae,
Micrococcaceae, Staphylococcaceae and Streptococcaceae and decreases in
Ruminococcaceae (6--9). Several of these taxa have also been associated
with \emph{C. difficile} colonization in humans (10).\\
Unfortunately, most of the studies suggesting a link between PPIs and
\emph{C. difficile} were retrospective and did not evaluate changes in
the microbiome (2, 3, 5). Thus, it is unclear whether the
gastrointestinal microbiome changes associated with PPI use explain the
association between PPIs and CDIs. Additionally, epidemiological studies
have a limited capacity to address potential confounders and
comorbidities in patients that were on PPIs and developed CDIs or
recurrent CDIs (2, 5). Here, we evaluated the impact of daily PPI
treatment with omeprazole on the murine microbiome and susceptibility to
\emph{C. difficile} colonization in relation to clindamycin, an
antibiotic that perturbs the microbiome enough to allow \emph{C.
difficile} to colonize but is mild enough that \emph{C. difficile} is
cleared within 10 days (11).

\textbf{Murine fecal microbiomes were minimally affected by omeprazole
treatment } To test whether omeprazole treatment alters the microbiome
and promotes susceptibility to CDIs, we gavaged mice with 40 mg/kg of
omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A).
A principle coordinates analysis (PCoA) of the Bray-Curtis distances
over the initial 7 days of treatment revealed the bacterial communities
of omeprazole-treated mice remained relatively unchanged (Figure 1B). We
observed no significant fluctuations in the relative abundance of those
taxa previously shown to respond to PPI treatment throughout the course
of the 16-day experiment (Figure 1C-D, S1). We also observed no
significant changes in relative abundances at the family and genus level
over the course of the experiment for the omeprazole-treated mice (data
not shown). These results demonstrated that the omeprazole treatment
alone had a minimal impact on the murine fecal bacterial community after
7 days of pretreatment.

\textbf{Omeprazole treatment did not promote susceptibility to \emph{C.
difficile infection} in mice} Next, we examined whether omeprazole
treatment altered susceptibility to \emph{C. difficile} infection in
mice. After omeprazole treatment or clindamycin treatment, mice were
challenged with 10\textsuperscript{3} \emph{C. difficile} 630 spores.
While \emph{C. difficile} colonized the clindamycin-treated mice, it did
not colonize the omeprazole-treated mice (Figure 2A). Interestingly,
only 1 cage of mice that received both omeprazole and clindamycin were
colonized, while the other cage of mice were resistant (Figure 2A). The
greatest shifts in bacterial communities occurred in the
clindamycin-treated mice (Figure 2B, S2). Regardless of whether the mice
became colonized, all of the mice had cleared \emph{C. difficile} within
5 days (Figure 2A), suggesting that omeprazole did not affect the rate
of clearance. Our results suggest that omeprazole treatment had no
effect on bacterial community resistance to \emph{C. difficile}
colonization in mice. Instead most of the differences between the 3
treatment groups appeared to be driven by clindamycin administration
(Figure 2C, S2). These findings demonstrated that high dose omeprazole
treatment did not promote susceptibility to \emph{C. difficile}
colonization.

\subsection{Conclusions}\label{conclusions}

We found the PPI omeprazole did not impact the gut microbiota and did
not promote \emph{C. difficile} infection in mice. Our findings that
omeprazole treatment had minimal impact on the fecal microbiome were
comparable to another PPI mouse study that indicated the PPI
lansoprazole had more of an effect on the small intestinal microbiota
compared to the fecal microbiota (12). The same group demonstrated
lansoprazole treatment increased the stomach pH in mice (12), which may
improve survival of bacteria passing through the stomach. We did not
find significant changes for the taxa observed to be significantly
impacted by PPI use in human studies. However, 3 of the human-associated
taxa were absent or at low abundance in our mice. Although our fecal
microbiota findings are comparable to what has been shown in another
mouse study (12), whether PPI-induced changes in specific bacterial
abundances observed in humans play a role in CDIs remains to be
determined.\\
While several \emph{C. difficile} mouse model studies have shown that

@@

antibiotic/PPI (esomeprazole)-treated mice developed more severe CDIs
compared to antibiotic-treated mice, but did not have a group treated
with just esomeprazole for comparison (15). We tested the same high 40
mg/kg PPI dose and expanded pre-treatment to 7 days before challenge to
test the impact of omeprazole treatment alone on our CDI mouse model.
Additionally, we've previously demonstrated that mice from our colony
were resistant to \emph{C. difficile} 630 colonization without
antibiotic treatment (16), ensuring there was not already partial
susceptiblity to \emph{C. difficile} before treatment. The additional
controls in our study allowed us to assess the contibution of omeprazole
alone to \emph{C. difficile} susceptibility in mice.\\

@@

microbiota. We found the PPI omeprazole had no significant impact on
bacterial taxa within the murine intestinal microbiota over the 16-day
experiment. In contrast to previous work, omeprazole did not alter
\emph{C. difficile} colonization resistance. 16S rRNA sequencing
suggested that \emph{Streptococcus} and \emph{Enterococcus} are rare
genera in our C57BL/6 mouse colony and variation in these 2 genera have
been observed across other facilities and vendors (17, 18). These genera
could be important contributors to the associations between PPIs and
CDIs in humans, and could be a contributing factor to our observation
that PPI treatment had no effect on \emph{C. difficile} colonization in
our CDI mouse model. Gastrointestinal physiological differences,
particularly the higher stomach pH in mice (pH 3-4) compared to humans
(ph 1) (19) could also explain why omeprazole had a limited impact on
the murine microbiome. The microbiota and physiological differences
between humans and mice may limit the usefulness of employing mouse
models to study the impact of PPIs on the microbiota and CDIs.\\

@@

\textbf{Statistical Analysis.} To test for differences in relative
abundances in families and genera across our 3 different treatment
groups (Clindamycin, Clindamycin + PPI, and PPI) or within the PPI
treatment group across 3 timepoints (Day -7, 0, and 9), we used a
Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple
comparisons.

\textbf{Code availability.} The code for all sequence processing and
analysis step as well as an Rmarkdown version of this manuscript is
available at
\url{https://github.com/SchlossLab/Tomkovich_PPI_mSphere_2019}.

@@

\includegraphics{figure_1.pdf} \textbf{Figure 1. Omeprazole treatment
had minimal impact on the murine fecal microbiota.} A. Mouse experiment
timeline and logistics. The PPI omeprazole was administered throughout
the duration of the experiment. Clindamycin was administered 1 day
before \emph{C. difficile} challenge on Day 0. Stools for 16S rRNA
sequencing analysis were collected on the days that are labeled (Day -7,
-5, -3, -1, 0, 1, 2, 3, 4, 5, 7, 9). \emph{C. difficile} CFU in the
stool was quantified daily through 6 days post-infection by anaerobic
culture. B. Principal Coordinates Analaysis (PCoA) of Bray-curtis
distances from stool samples of mice in the omeprazole treatment group
during the intial 7 days of the experiment. Each color represents stool
samples from the same mouse and lines connect sequentially collected
samples. C-D. Relative abundances of families previously associated with
PPI use in humans at the start of the experiment (C) and after 7 days of
omeprazole treatment (D). Each circle represents an individual mouse.
There were no significant differences acoss treatment groups for any of
the identified families in the seqeunce data at day -7 and day 0,
analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for
multiple comparisons. For C-D, the grey vertical line indicates the
limit of detection.

@@

\includegraphics{figure_2.pdf} \textbf{Figure 2. Omeprazole treatment
alone does not promote CDIs in mice.} A. \emph{C. difficile} CFUs/g
stool measured each day post \emph{C. difficile} challenge for
clindamycin, clindamycin/PPI, and PPI-treated mice. Lines represent the
mean CFU/g for each treatment group while points represent CFU/g for
individual mice within each group. The black dashed line indicates the
limit of detection. B. PCoA of of Bray-curtis distances from stool
samples collected after antibiotic treatment (last 9 days of the
experiment). Transparency of the symbol corresponds to treatment day.
Symbols represent the \emph{C. difficile} colonization status of the
mice measured 2 days post-infection. Circles represent resistant mice
(\emph{C. difficile} was undetectable in stool samples), while X-shapes
represent mice that were colonized with \emph{C. difficile}, although
all mice cleared \emph{C. difficile} within 5 days of infection.
Omeprazole treated fecal samples primarily cluster together throughout
the experiment. C. Genera that vary the most across treatment groups for
stool samples collected from mice 2 days post-infection. Data were
analyzed by Kruskal-Wallis test, and no P values were significant after
Benjamini-Hochberg correction for multiple comparisons. The grey
vertical line indicates the limit of detection.

@@

\includegraphics{figure_s1.pdf} \textbf{Figure S1. Families within
omeprazole treated mice fluctuate over time with no overall trend in
either direction.} Relative abundance over time for Lactobacillaceae (A)
and Ruminococcaceae (B), 2 of the PPI-associated familes from human PPI
studies across all 3 treatment groups. Each point represents the
relative abundance for an individual mouse stool sample, while the lines
represent the mean relative abundances for each treatment group of mice.
The grey horizontal lines indicate the limit of detection.

>>>>

Proton pump inhibitor (PPI) use has been associated with microbiota
alterations and susceptibility to \emph{Clostridium difficile}
infections (CDIs) in humans. We assessed how PPI treatment alters the
fecal microbiota and whether treatment promotes CDIs in a mouse model.
Mice receiving a PPI treatment were gavaged with 40 mg/kg of omeprazole
during a 7-day pretreatment phase, the day of \emph{C. difficile}
challenge, and the following 9 days. We found that mice treated with
omeprazole were not colonized by \emph{C. difficile}. When omeprazole
treatment was combined with a single clindamycin treatment, one cage of
mice remained resistant to \emph{C. difficile} colonization, while the
other cage was colonized. Treating mice with only clindamycin followed
by challenge resulted in \emph{C. difficile} colonization. 16S rRNA gene
sequencing analysis revealed that omeprazole had minimal impact on the
structure of the murine microbiota throughout the 16 days of omeprazole
exposure. These results suggest omeprazole treatment alone is not
sufficient to disrupt microbiota resistance to \emph{C. difficile}
infection in mice that are normally resistant in the absence of
antibiotic treatment.

@@

Antibiotics are the primary risk factor for \emph{Clostridium difficile}
infections (CDIs), but other factors may also increase a person's risk.
In epidemiological studies, proton pump inhibitor (PPI) use has been
associated with CDI incidence and recurrence. PPIs have also been
associated with alterations in the human intestinal microbiota in
observational and interventional studies. We evaluated the effects of
the PPI omeprazole on the structure of the murine intestinal microbiota
and its ability to disrupt colonization resistance to \emph{C.
difficile}. We found omeprazole treatment had minimal impact on the
murine fecal microbiota and did not promote \emph{C. difficile}
colonization. Further studies are needed to determine whether other
factors such as the composition of the starting bacterial community,
comorbidities, and use of additional medications contribute to the
association between PPIs and CDIs seen in humans or whether aspects of
murine physiology may limit its utility to test these types of
hypotheses.

@@

Antibiotics have a large impact on the intestinal microbiome and are a
primary risk factor for developing \emph{Clostridium difficile}
infections (CDIs) (1). It is less clear whether other human medications
that impact the microbiota also influence \emph{C. difficile}
colonization resistance. Multiple epidemiological studies have suggested
an association between proton pump inhibitor (PPI) use and incidence or
recurrence of CDIs (2--5). There have also been a number of large cohort
studies and interventional clinical trials that demonstrated specific
alterations in the intestinal microbiome were associated with PPI use
(4, 6). PPI-associated microbiota changes have been attributed to the
ability of PPIs to increase stomach acid pH which may promote the
survival of oral and pathogenic bacteria (4, 6). In human fecal samples,
PPI use results in increases in \emph{Enterococcaceae},
\emph{Lactobacillaceae}, \emph{Micrococcaceae}, \emph{Staphylococcaceae}
and \emph{Streptococcaceae} and decreases in \emph{Ruminococcaceae}
(6--9). Several of these taxa have also been associated with \emph{C.
difficile} colonization in humans (10).\\
Unfortunately, the studies suggesting a link between PPIs and \emph{C.
difficile} were retrospective and did not evaluate changes in the
microbiome (2, 3, 5). Thus, it is unclear whether the gastrointestinal
microbiome changes associated with PPI use explain the association
between PPIs and CDIs. Additionally, epidemiological studies have a
limited capacity to address potential confounders and comorbidities in
patients that were on PPIs and developed CDIs or recurrent CDIs (2, 5).
Here, we evaluated the impact of daily PPI treatment with omeprazole on
the murine microbiome and susceptibility to \emph{C. difficile}
colonization in relation to clindamycin, an antibiotic that perturbs the
microbiome enough to allow \emph{C. difficile} to colonize but is mild
enough that \emph{C. difficile} is cleared within 10 days (11).

\textbf{Murine fecal microbiomes were minimally affected by omeprazole
treatment. } To test whether omeprazole treatment alters the microbiome
and promotes susceptibility to CDIs, we gavaged mice with 40 mg/kg of
omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A).
A principle coordinates analysis (PCoA) of the Bray-Curtis distances
over the initial 7 days of treatment revealed the bacterial communities
of omeprazole-treated mice remained relatively unchanged (Figure 1B). We
observed no significant changes in the relative abundance of those taxa
previously shown to respond to PPI treatment throughout the course of
the 16-day experiment (Figure 1C-D, S1). We also observed no significant
changes in relative abundances at the family and genus level over the
course of the experiment for the omeprazole-treated mice (all corrected
P-values \textgreater{} 0.36). These results demonstrated that the
omeprazole treatment alone had a minimal impact on the murine fecal
bacterial community after 7 days of pretreatment.

\textbf{Omeprazole treatment did not promote susceptibility to \emph{C.
difficile infection} in mice.} Next, we examined whether omeprazole
treatment altered susceptibility to \emph{C. difficile} infection in
mice. After omeprazole treatment or clindamycin treatment, mice were
challenged with 10\textsuperscript{3} \emph{C. difficile} 630 spores.
While \emph{C. difficile} colonized the clindamycin-treated mice, it did
not colonize the omeprazole-treated mice (Figure 2A). Interestingly,
only 1 cage of mice that received both omeprazole and clindamycin were
colonized, while the other cage of mice were resistant (Figure 2A). The
greatest shifts in bacterial communities occurred in the
clindamycin-treated mice (Figure 2B, S2). Regardless of whether the mice
became colonized, all of the mice had cleared \emph{C. difficile} within
5 days (Figure 2A), suggesting that omeprazole did not affect the rate
of clearance. Our results suggest that omeprazole treatment had no
effect on bacterial community resistance to \emph{C. difficile}
colonization in mice. Instead most of the differences between the 3
treatment groups appeared to be driven by clindamycin administration
(Figure 2C, S2). These findings demonstrated that high dose omeprazole
treatment did not promote susceptibility to \emph{C. difficile}
colonization.

\textbf{Conclusions.} The PPI omeprazole did not meaningfully impact the
gut microbiota and did not promote \emph{C. difficile} infection in
mice. Our findings that omeprazole treatment had minimal impact on the
fecal microbiome were comparable to another PPI mouse study that
indicated the PPI lansoprazole had more of an effect on the small
intestinal microbiota compared to the fecal microbiota (12). The same
group demonstrated lansoprazole treatment increased the stomach pH in
mice (12), which may improve survival of bacteria passing through the
stomach. We did not find significant changes in the relative abundances
of the taxa observed to be significantly impacted by PPI use in human
studies. However, 3 of the human-associated taxa were absent or at low
abundance in our mice. Although our fecal microbiota findings are
comparable to what has been shown in another mouse study (12), whether
PPI-induced changes in specific bacterial abundances observed in humans
play a role in CDIs remains to be determined.\\
Although several \emph{C. difficile} mouse model studies have shown that

@@

antibiotic/PPI esomeprazole-treated mice developed more severe CDIs
compared to antibiotic-treated mice, but the researchers did not have a
group treated with just esomeprazole for comparison (15). We tested the
same high 40 mg/kg PPI dose and expanded pre-treatment to 7 days before
challenge to test the impact of omeprazole treatment alone on our CDI
mouse model. Additionally, we have previously demonstrated that mice
from our breeding colony are resistant to \emph{C. difficile} 630
colonization without antibiotic treatment (16), ensuring there was not
already partial susceptiblity to \emph{C. difficile} before treatment.
The additional controls in our study allowed us to assess the
contibution of omeprazole alone to \emph{C. difficile} susceptibility in
mice.\\

@@

microbiota. We found omeprazole had no significant impact on bacterial
taxa within the murine intestinal microbiota over the 16-day experiment.
In contrast to previous work with PPIs (13--15), omeprazole did not
alter \emph{C. difficile} colonization resistance in mice. 16S rRNA
sequencing suggested that \emph{Streptococcus} and \emph{Enterococcus}
are rare genera in our C57BL/6 mouse colony and variation in these 2
genera have been observed across other facilities and vendors (17, 18).
These genera could be important contributors to the associations between
PPIs and CDIs in humans, and could be a contributing factor to our
observation that PPI treatment had no effect on \emph{C. difficile}
colonization in our CDI mouse model. Gastrointestinal physiological
differences, particularly the higher stomach pH in mice (pH 3-4)
compared to humans (pH 1) (19) could also explain why omeprazole had a
limited impact on the murine microbiome. The microbiota and
physiological differences between humans and mice may limit the
usefulness of employing mouse models to study the impact of PPIs on the
microbiota and CDIs.\\

@@

\textbf{Statistical Analysis.} To test for differences in relative
abundances in families and genera across our 3 different treatment
groups at different timepoints (Clindamycin, Clind. + Omep., and
Omeprazole on Day -7, 0, 2, and 9) or within the Omeprazole treatment
group across 3 timepoints (Day -7, 0, and 9), we used a Kruskal-Wallis
test with a Benjamini-Hochberg correction for multiple comparisons.

\textbf{Code availability.} The code for all sequence processing and
analysis steps as well as an Rmarkdown version of this manuscript is
available at
\url{https://github.com/SchlossLab/Tomkovich_PPI_mSphere_2019}.

@@

\includegraphics{figure_1.pdf} \textbf{Figure 1. Omeprazole treatment
had minimal impact on the murine fecal microbiota.} A. Mouse experiment
timeline and logistics. The PPI omeprazole was administered throughout
the duration of the experiment. Clindamycin was administered 1 day
before \emph{C. difficile} challenge on Day 0. Stools for 16S rRNA
sequencing analysis were collected on the days that are labeled (Day -7,
-5, -3, -1, 0, 1, 2, 3, 4, 5, 7, 9). \emph{C. difficile} CFU in the
stool was quantified daily through 6 days post-infection by anaerobic
culture. B. Principal Coordinates Analaysis (PCoA) of Bray-curtis
distances from stool samples of mice in the omeprazole treatment group
during the intial 7 days of the experiment. Each color represents stool
samples from the same mouse and lines connect sequentially collected
samples. C-D. Relative abundances of families previously associated with
PPI use in humans at the start of the experiment (C) and after 7 days of
omeprazole treatment (D). Each circle represents an individual mouse.
There were no significant differences acoss treatment groups for any of
the identified families in the sequence data at day -7 (all P-values
\textgreater{} 0.448) and day 0 (all P-values \textgreater{} 0.137),
analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for
multiple comparisons. For C-D, the grey vertical line indicates the
limit of detection.

@@

\includegraphics{figure_2.pdf} \textbf{Figure 2. Omeprazole treatment
alone does not promote CDIs in mice.} A. \emph{C. difficile} CFUs/g
stool measured each day post \emph{C. difficile} challenge for
clindamycin, clindamycin/omeprazole, and omeprazole-treated mice. Lines
represent the mean CFU/g for each treatment group while points represent
CFU/g for individual mice within each group. The black dashed line
indicates the limit of detection. B. PCoA of of Bray-curtis distances
from stool samples collected after antibiotic treatment (last 9 days of
the experiment). Transparency of the symbol corresponds to treatment
day. Symbols represent the \emph{C. difficile} colonization status of
the mice measured 2 days post-infection. Circles represent resistant
mice (\emph{C. difficile} was undetectable in stool samples), while
X-shapes represent mice that were colonized with \emph{C. difficile},
although all mice cleared \emph{C. difficile} within 5 days of
infection. Omeprazole treated fecal samples primarily cluster together
throughout the experiment. C. Genera that vary the most across treatment
groups for stool samples collected from mice 2 days post-infection. Data
were analyzed by Kruskal-Wallis test, and no P-values were significant
after Benjamini-Hochberg correction for multiple comparisons (all
P-values \textgreater{} 0.092). The grey vertical line indicates the
limit of detection.

@@

\includegraphics{figure_s1.pdf} \textbf{Figure S1. Families within
omeprazole treated mice fluctuate over time with no overall trend in
either direction.} Relative abundance over time for
\emph{Lactobacillaceae} (A) and \emph{Ruminococcaceae} (B), 2 of the
PPI-associated familes from human PPI studies across all 3 treatment
groups. Each point represents the relative abundance for an individual
mouse stool sample, while the lines represent the mean relative
abundances for each treatment group of mice. The grey horizontal lines
indicate the limit of detection.
